---
title: "PS270 Final Project"
author: "Ellie DeCleene"
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---

```{r include=FALSE}
library(tidyverse)
library(gssr)
```

## Project thoughts

I am interested in exploring data related to...
 
- How do people's political beliefs change over time? Are people more likely to agree with Democrats when they're young and change their minds as they age? Or do people generally stay in the same party as they age, and the Republican share of the vote increases with age because previous non-voters begin to vote?

- How well do your parents' beliefs predict your beliefs?

- Is the number of different news sources consumed by a voter predictive of their political party?

## Project plan

How are people's parents impact their political views? My hypothesis is that people who have less success than their parents are more likely to believe that the country is on the wrong track and be more conservative, while people who have more success than their parents are more likely to believe that the country is on the right track and be more progressive. My explanatory variable would be whether the individual has more success than their parents, measured by income, occupation, education; my outcome variable would be the response to questions such as "Is the country on the right track?" and overall political party affiliation. I would potentially want to control for income level, because I am curious more about comparative income than absolute income. My hypothesis would be supported if there is a positive correlation between success relative to your parents and progressive beliefs; my hypothesis would be disproved if there is no relation or if there is a negative correlation.

## Loading data

```{r}
gss_2018 = gss_get_yr(year = 2018) %>% 
  select(prestg10, sppres10, papres10, mapres10, codeg, degree, padeg, madeg, spdeg, res16, incom16, income, rincome, income16, pres16, polviews, happy, life, helpful, fair, trust, satjob, rank, satfin, finrela, parsol, unhappy, cntctpar)
```

## Initial plot

```{r fig.width=9, fig.height=7}
polviews_vs_diff = gss_2018 %>% group_by(polviews) %>% summarize(me=mean(income, na.rm=TRUE), par=mean(incom16, na.rm=TRUE), diff=me-par) %>% head(7)

polviews_vs_diff %>% ggplot(mapping=aes(x=polviews, y=diff)) +
  geom_point() +
  geom_smooth(method="lm", se=FALSE) +
  ggtitle("How Are Your Political Views Related to\nYour Income vs. Your Parents' Income?") +
  labs(x="Political Views", y="Present Income minus Childhood Family Income") +
  scale_x_continuous(breaks=c(2, 6), labels=c("More Conservative", "More Liberal")) +
  scale_y_continuous(breaks=c(8.1, 8.5), labels=c("   Less Income\nthan Parents", "   More Income\nthan Parents")) +
  theme_light() +
  theme(
    axis.text.x = element_text(size=18), 
    axis.title.x = element_text(size=18),
    text = element_text(size=18)
  ) +
  annotate("text", x=6.4, y=8, label="Sourced from the General\nSocial Survey (2018)")
```

This plot suggests that there is a positive correlation between liberal beliefs and income relative to your parents'. 

```{r}
self_vs_diff = gss_2018 %>% 
  mutate(`Income Relative to Parents` = income-incom16, `Income` = income) %>% 
  lm(formula=polviews ~ `Income Relative to Parents` + `Income`) 

summary(self_vs_diff)$coefficients %>% 
  knitr::kable()
```
This linear model provides further evidence for the relationship. When considering both income relative to parents, and income without comparison, we can see that income relative to parents is much more statistically significant in predicting political views than income itself ($p < 0.1$ vs $p > 0.3$). 

These findings support my hypothesis. There is a positive correlation between success (measured as income) compared to your parents and progressive policy affiliation. It cannot be definitively concluded that this is related to an underlying variable of whether one believes the country is on the right/wrong track, but that would be a reasonable way to potentially explain the results.
