---
title: "PS270 Final Project"
author: "Ellie DeCleene"
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---

```{r include=FALSE}
library(tidyverse)
library(gssr)
```

## Project thoughts

I am interested in exploring data related to...
 
- How do people's political beliefs change over time? Are people more likely to agree with Democrats when they're young and change their minds as they age? Or do people generally stay in the same party as they age, and the Republican share of the vote increases with age because previous non-voters begin to vote?

- How well do your parents' beliefs predict your beliefs?

- Is the number of different news sources consumed by a voter predictive of their political party?

## Project plan

How are people's parents impact their political views? My hypothesis is that people who have less success than their parents are more likely to believe that the country is on the wrong track and be more conservative, while people who have more success than their parents are more likely to believe that the country is on the right track and be more progressive. My explanatory variable would be whether the individual has more success than their parents, measured by income, occupation, education; my outcome variable would be the response to questions such as "Is the country on the right track?" and overall political party affiliation. I would potentially want to control for income level, because I am curious more about comparative income than absolute income. My hypothesis would be supported if there is a positive correlation between success relative to your parents and progressive beliefs; my hypothesis would be disproved if there is no relation or if there is a negative correlation.

## Loading data

```{r}
gss_2018 = gss_get_yr(year = 2018) %>% 
  select(prestg10, sppres10, papres10, mapres10, codeg, degree, padeg, madeg, spdeg, res16, incom16, income, rincome, income16, pres16, polviews, happy, life, helpful, fair, trust, satjob, rank, satfin, finrela, parsol, unhappy, cntctpar)
```

## Initial plot

```{r fig.width=10, fig.height=8}
polviews_vs_diff = gss_2018 %>% group_by(polviews) %>% summarize(me=mean(income, na.rm=TRUE), par=mean(incom16, na.rm=TRUE), diff=me-par) %>% head(7)

polviews_vs_diff %>% ggplot(mapping=aes(x=polviews, y=diff)) +
  geom_point() +
  geom_smooth(method="lm", se=FALSE) +
  labs(x="Political Views", y="Present Income minus Childhood Family Income", title="How Are Your Political Views Related to Your Income vs. Your Parents' Income?") +
  scale_x_continuous(breaks=c(2, 6), labels=c("More Conservative", "More Liberal")) +
  scale_y_continuous(breaks=c(8.1, 8.5), labels=c("Higher Income in Childhood", "Higher Income Now"))

```